Holomorphic field theories and higher algebra

### 2023-08-18 highervertexops.tex
more polishing of the sl2 features of A

Let $\sfR$ be the commutative algebra generated by variables $z_i,\lambda_i,i=1,2$ subject to the relation $\sum_i z_i \lambda_i = 1$.
Let $\sfA$ be the graded commutative algebra freely generated over $\sfR$ by a degree $+1$ element $\omega$.
Thus $\sfA = \sfR \oplus \sfR \omega [-1]$.
Equip it with the differential
\[
\dbar(z_i) = 0 \quad\text{and}\quad \dbar(\lambda_i) = \epsilon_{ij} z_j \omega
\]
extended as a derivation.
That is, $\dbar(\lambda_1) = z_2 \omega$ and $\dbar(\lambda_1) = -z_1 \omega$.

@@

Recall that $\Sym^a(V_z)$ is an irreducible representation of dimension $a+1$ (and hence highest weight $a$) and $\Sym^b(V_\lambda)$ is an irreducible representation of dimension $b+1$ (and hence highest weight $a$).
Let $I_m$ denote the irreducible representation of highest weight $m$ (and hence of dimension $m+1$).
By standard facts about $SL(2)$, we see that $V_{a,b}$ decomposes as a direct sum of irreducible representations
\[
V_{a,b} \cong I_{a+b} \oplus I_{a+b-2} \oplus \cdots \oplus I_{a-b}
\]
for $a \geq b$. (The other situation can be seen by symmetry.)

There is a more concrete way to describe this decomposition, 
modeled on the construction of spherical harmonics as representations of $SO(n)$.
Consider the differential operator
\beqn
\triangle = \del_{z_1} \del_{\lambda_1} + \del_{z_2} \del_{\lambda_2}
\eeqn
acting on $\C[z_i,\lambda_i]$;
it is inspired by the Laplacian on $\CC^2$ when we view $\lambda_i$ as a version of $\Bar{z_i}$.
This operator has weight zero and, in fact, it is $SL(2)$-equivariant.
For example, notice that for any $g(z_i,\lambda_i)$ we have
\begin{align*}
(\del_{z_1} \del_{\lambda_1} + \del_{z_2} \del_{\lambda_2})(z_2 \del_{z_1} g - \lambda_1 \del_{\lambda_2}g )& = z_2 \del_{z_1}^2 \del_{\lambda_1} g - \del_{z_1} \del_{\lambda_2} g - \lambda_1 \del_{z_1} \del_{\lambda_1} \del_{\lambda_2} g \\ & \qquad + \del_{z_1} \del_{\lambda_2}  g + z_2 \del_{z_1} \del_{z_2} \del_{\lambda_2} g - \lambda_1 \del_{z_2} \del_{\lambda_2}^2 g 
\\ & = z_2 \del_{z_1}^2 \del_{\lambda_1} g - \lambda_1 \del_{z_1} \del_{\lambda_1} \del_{\lambda_2} g + z_2 \del_{z_1} \del_{z_2} \del_{\lambda_2} g - \lambda_1 \del_{z_2} \del_{\lambda_2}^2 g .
\end{align*}
Thus $[\triangle,\sff] = 0$.
Similarly, one can show that $[\triangle, \sfe] = 0$.
\owen{I need to check this myself. If there's a slick way, we should include it.} \brian{I gave a computational proof.}
On the other hand, it maps $V_{a,b}$ to $V_{a-1,b-1}$, i.e., it changes the bigrading by~$(-1,-1)$.

Let $H_{a,b} = \ker \triangle|_{V_{a,b}}$ denote the ``harmonic'' polynomials of bidegree $(a,b)$.
Examples of harmonic polynomials include $z_1^a \lambda_2^b$ for arbitrary $a,b$ and
$z_1 \lambda_1 - z_2 \lambda_2$ for $a = 1 = b$.
A direct computation shows that there is a direct sum decomposition \owen{of representations?}
\begin{align*}
V_{a,b} & = H_{a,b} \oplus |(z,\lambda)|^2 V_{a-1,b-1} \\
& = H_{a,b} \oplus |(z,\lambda)|^2 H_{a-1,b-1} \oplus |(z,\lambda)|^4 H_{a-2,b-2} \oplus \cdots 
\end{align*}
(Compare with the well-known )

Consider now  
\[
|(z,\lambda)|^2 \define z_1 \lambda_1 + z_2 \lambda_2, 
\]
which describes the evaluation pairing. 
It is preserved by the $SL(2)$-action, and hence we see that the canonical quotient map 
\[
q \colon \C[z_i,\lambda_i] \to \sfR = \C[z_i,\lambda_i]/(|(z,\lambda)|^2 = 1)
\]
is $SL(2)$-equivariant.
Moreover, inside $V_{a,b}$, any multiple of $|(z,\lambda)|^2$ is annihilated by $q$,
so $R$ is spanned by the images of $H_{a,b}$.
Thus, there is a direct sum decomposition 
\beqn
\sfR = \bigoplus_{a,b} \sfR_{a,b}
\eeqn
where $\sfR_{a,b} = q(H_{a,b})$ is the image of the subspace of harmonic polynomials in $\C[z_i,\lambda_i]$ under the quotient map.
We remark that this decomposition does \textit{not} respect the commutative product.

@@

\begin{cor}
The differential $\dbar$ is an isomorphism from $\sfR_{a,b}$ to $\sfR_{a+1,b-1} \omega$ for $b > 0$.
Thus, the cohomology of $\sfA$ is 
\[
H^0(\sfA) = \bigoplus_{a \in \NN} \sfR_{a,0}
\]
and
\[
H^1(\sfA) = \bigoplus_{b \in \NN} \sfR_{0,b}.
\]
\end{cor}

@@

There is a canonical contracting homotopy $H$ on $\sfA$ 
because for each $b > 0$, there is a canonical equivariant inverse $\sfR_{a,b} \to \sfR_{a+1,b-1} \omega$.
Explicitly, send the highest weight vector $z_1^{a+1} \lambda_2^{b-1} \omega$ to the highest weight vector $z_1^{a} \lambda_2^{b}$ and then determine where other vector go via the lowering operator.
For $b = 0$, $H|_{\sfR_{0,b}\omega} = 0$.

@@

Note that $p$ is not a map of algebras. 
For example,
\beqn
p(z_1 \lambda_1) = \frac12
\eeqn
as $z_1 \lambda_1$ admits a $\lie{sl}(2)$-decomposition
\beqn
z_1 \lambda_1 =  \frac12  + \frac12 (z_1 \lambda_1 - z_2 \lambda_2).
\eeqn
\subsection{The higher algebraic residue}

@@

\begin{lem}
\label{lem:res} 
The following are true regarding the higher algebraic residue:
\begin{enumerate}
\item The residue is $\lie{sl}(2)$ invariant.
\item For $f \in \C[[z_1,z_2]]$ one has $\op{Res}(f \omega) = f(0)$.
\item For any $f \in \laurent$ one has $\op{Res}(\dbar f) = 0$.
Thus $\op{Res}$ is a cochain map.
\end{enumerate}
\end{lem}
\begin{proof}
Item (1) holds since $p$ is the $\lie{sl}(2)$-equivariant projection. 
For (2), note that for $f \in \C[z_1,z_2]$ that $p(f) = f$.

>>>>

Let $\sfR$ be the commutative algebra generated by variables $z_i,\lambda_i,i=1,2$ subject to the relation $z_1\lambda_1 + z_2 \lambda_2 = 1$.
Let $\sfA$ be the graded commutative algebra freely generated over $\sfR$ by a degree $+1$ element $\omega$.
Thus $\sfA = \sfR \oplus \sfR \omega [-1]$.
Equip it with the differential
\[
\dbar(z_i) = 0 \quad\text{and}\quad \dbar(\lambda_i) = \epsilon_{ij} z_j \omega
\]
extended as a derivation.
That is, $\dbar(\lambda_1) = z_2 \omega$ and $\dbar(\lambda_1) = -z_1 \omega$.

@@

Recall that $\Sym^a(V_z)$ is an irreducible representation of dimension $a+1$ (and hence highest weight $a$) and $\Sym^b(V_\lambda)$ is an irreducible representation of dimension $b+1$ (and hence highest weight $a$).
Let $I_m$ denote the irreducible representation of highest weight $m$ (and hence of dimension $m+1$).
By standard facts about $SL(2)$, we see that $V_{a,b}$ decomposes as a direct sum of irreducible representations
\begin{equation}
\label{eq:sl2decompV}
V_{a,b} \cong I_{a+b} \oplus I_{a+b-2} \oplus \cdots \oplus I_{a-b}
\end{equation}
for $a \geq b$. (The other situation can be seen by symmetry.)

There is a more concrete way to describe this decomposition, 
modeled on the construction of spherical harmonics as representations of $SO(n)$.
Consider the element  
\[
|(z,\lambda)|^2 \define z_1 \lambda_1 + z_2 \lambda_2, 
\]
which describes the evaluation pairing. 
It is fixed by the $SL(2)$-action, has weight zero, and bigrading $(1,1)$.
Multiplication by this element produces an injection
\[
V_{a-1,b-1} \hookrightarrow V_{a,b}
\]
that is $SL(2)$-equivariant, so we know there is a decomposition 
\[
V_{a,b} \cong |(z,\lambda)|^2 V_{a-1,b-1} \oplus C_{a,b},
\]
where $C_{a,b}$ is a complementary subspace and representation.
We can characterize this subrepresentation concretely using the differential operator
\beqn
\triangle = \del_{z_1} \del_{\lambda_1} + \del_{z_2} \del_{\lambda_2}
\eeqn
acting on $\C[z_i,\lambda_i]$;
it is inspired by the Laplacian on $\CC^2$ when we view $\lambda_i$ as a version of $\Bar{z_i}$.
This operator has weight zero and, in fact, it is $SL(2)$-equivariant.\footnote{For example, notice that for any $g(z_i,\lambda_i)$ we have
\begin{align*}
\triangle \circ \sff(g) &= (\del_{z_1} \del_{\lambda_1} + \del_{z_2} \del_{\lambda_2})(z_2 \del_{z_1} g - \lambda_1 \del_{\lambda_2}g )\\
&= z_2 \del_{z_1}^2 \del_{\lambda_1} g - \del_{z_1} \del_{\lambda_2} g - \lambda_1 \del_{z_1} \del_{\lambda_1} \del_{\lambda_2} g \\ & \qquad + \del_{z_1} \del_{\lambda_2}  g + z_2 \del_{z_1} \del_{z_2} \del_{\lambda_2} g - \lambda_1 \del_{z_2} \del_{\lambda_2}^2 g \\ 
& = z_2 \del_{z_1}^2 \del_{\lambda_1} g - \lambda_1 \del_{z_1} \del_{\lambda_1} \del_{\lambda_2} g + z_2 \del_{z_1} \del_{z_2} \del_{\lambda_2} g - \lambda_1 \del_{z_2} \del_{\lambda_2}^2 g \\
&=  \sff \circ \triangle(g)
\end{align*}
Thus $[\triangle,\sff] = 0$.
Similarly, one can show that $[\triangle, \sfe] = 0$.)
\owen{I need to check this myself. If there's a slick way, we should include it.} \brian{I gave a computational proof.}\owen{Thank you! I don't know why I'm so lazy about calculus.}}
On the other hand, it maps $V_{a,b}$ to $V_{a-1,b-1}$, i.e., it changes the bigrading by~$(-1,-1)$.

Let $H_{a,b} = \ker \triangle|_{V_{a,b}}$ denote the ``harmonic'' polynomials of bidegree $(a,b)$.
Examples of harmonic polynomials include $z_1^a \lambda_2^b$ for arbitrary $a,b$ and
$z_1 \lambda_1 - z_2 \lambda_2$ for $a = 1 = b$.
Direct computation shows that 
\[
V_{a,b}  = H_{a,b} \oplus |(z,\lambda)|^2 V_{a-1,b-1},
\]
and by iteratively applying this decomposition, we find
there is a direct sum decomposition
\[
V_{a,b}  = H_{a,b} \oplus |(z,\lambda)|^2 H_{a-1,b-1} \oplus |(z,\lambda)|^4 H_{a-2,b-2} \oplus \cdots \oplus |(z,\lambda)|^{2b} H_{a-b,0} 
\]
when $a \geq b$.
Thus $C_{a,b} = H_{a,b}$, and we also find $I_{a+b} = H_{a,b}$ by comparing with the decomposition~\eqref{eq:sl2decompV}.

In light of these facts, we see that the canonical quotient map 
\[
q \colon \C[z_i,\lambda_i] \to \sfR = \C[z_i,\lambda_i]/(|(z,\lambda)|^2 = 1)
\]
is $SL(2)$-equivariant.
Moreover, inside $V_{a,b}$, any multiple of $|(z,\lambda)|^2$ is annihilated by $q$,
so $\sfR$ is spanned by the images of $H_{a,b}$.
Thus, there is a direct sum decomposition 
\beqn
\sfR = \bigoplus_{a,b} \sfR_{a,b}
\eeqn
where $\sfR_{a,b} = q(H_{a,b})$ is the image of the subspace of harmonic polynomials in $\C[z_i,\lambda_i]$ under the quotient map.

We emphasize that this decomposition does \textit{not} respect the commutative product.

@@

\begin{cor}
The differential $\dbar$ provides an isomorphism 
\[
\dbar_{a,b} \colon \sfR_{a,b} \xto{\cong} \sfR_{a+1,b-1} \omega
\]
for $b > 0$.
Thus, the cohomology of $\sfA$ is 
\[
H^0(\sfA) = \bigoplus_{a \in \NN} \sfR_{a,0}
\]
and
\[
H^1(\sfA) = \bigoplus_{b \in \NN} \sfR_{0,b}.
\]
\end{cor}

@@

There is a canonical contracting homotopy $H$ on $\sfA$ 
because for each $b > 0$, there is a canonical equivariant inverse map $\dbar_{a,b}^{-1}:  \sfR_{a+1,b-1} \omega \to \sfR_{a,b}$.
Explicitly, send the highest weight vector $z_1^{a+1} \lambda_2^{b-1} \omega$ to the highest weight vector $z_1^{a} \lambda_2^{b}$ and then determine where other vector go via the lowering operator.
For $b = 0$, $H|_{\sfR_{0,b}\omega} = 0$.

@@

Note that $p$ is not a map of algebras. 
For example,
the following three polynomials in $\C[z_i,\lambda_i]$ represent the same element in~$\sfR$:
\beqn
\frac12  + \frac12 (z_1 \lambda_1 - z_2 \lambda_2) \sim \frac12  + \frac12 (z_1 \lambda_1 - z_2 \lambda_2) + \frac12 (z_1 \lambda_1 + z_2 \lambda_2) \sim z_1 \lambda_1
\eeqn
so
\beqn
p(z_1 \lambda_1) = p\left(\frac12  + \frac12 (z_1 \lambda_1 - z_2 \lambda_2)\right) = \frac12.
\eeqn

\subsection{The higher algebraic residue}

@@

\begin{lem}
\label{lem:res} 
The following are true regarding the higher algebraic residue:
\begin{enumerate}
\item The residue is $\lie{sl}(2)$-invariant.
\item For $f \in \C[[z_1,z_2]]$ one has $\op{Res}(f \omega) = f(0)$.
\item For any $f \in \laurent$ one has $\op{Res}(\dbar f) = 0$.
Thus $\op{Res}$ is a cochain map.
\end{enumerate}
\end{lem}
\begin{proof}
Item (1) holds since $p$ is the $\lie{sl}(2)$-equivariant projection. 
For (2), note that for $f \in \C[z_1,z_2]$ that $p(f) = f$.
